#Report Writing for Data Science in R Project 1
Zexia Hao  
<br> September 5th 2016

##Data
The data for this assignment can be downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken




The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this
dataset.

##Loading and preprocessing the data
###1. Load the data
```{r}

data<- read.csv("activity.csv")

```
###2. Transform the data into a format suitable using dplyr
```{r}
library(dplyr)
data<-tbl_df(data)
data
```
##What is the mean total number of steps taken per day?

###1.Calculate the total number of stpes taken per day without missing values
```{r}
#group data with date 
by_date<-group_by(data,date)
by_date
#calculate the total number,mean and medium of steps taken per day
stepsByDay<- summarize(by_date, total=sum(steps))
stepsByDay

```
###2.Calculate and report the mean and median of the total number of steps taken per day without missing values
```{r}
meanTotalStepsByDay<- mean(stepsByDay$total,na.rm=TRUE)
meanTotalStepsByDay

medianTotalStepsByDay<- median(stepsByDay$total,na.rm = TRUE)
medianTotalStepsByDay


```
###3.Make a histogram of the total number of steps taken per day
```{r}
with(stepsByDay, {    
  barplot(height=stepsByDay$total, names.arg=date, xlab="date", ylim=c(0,25000),
          ylab="# of steps", main = "Number of steps by day", panel.first=grid())
})
abline(h=meanTotalStepsByDay, col="blue")
abline(h=medianTotalStepsByDay, col="red")

```

##What is the average daily activity partern?

###1. Time series plot
```{r}
#construct a time series of the average number of steps taken per day and measured at each 5-minute interval
by_interval<-group_by(data,interval)
stepsByInterval<- summarize(by_interval,mean=mean(steps,na.rm=TRUE),total=sum(steps,na.rm = TRUE))
stepsByInterval

# make time series plot
with(stepsByInterval, {
  plot(stepsByInterval$interval, stepsByInterval$mean, type="l", xlab="intervals (hours)", ylab="mean of steps", 
       main="Average daily activity pattern", panel.first=grid(), 
       col.axis="white")
  
  # Add minor tick marks
  xMarks <- seq(from = 0, to = 2400, by = 100)
  yMarks <- seq(from = 0, to = 200, by = 50)
  axis(1, at=xMarks, col.axis="blue", las=2, tck=0.02)
  axis(2, at=yMarks, col.axis="blue", las=2, tck=0.02)
  
})

```

###2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
# maximum number of steps
maxSteps <- stepsByInterval$interval[which.max(stepsByInterval$total)]
maxSteps

```

##Imputing missing values

###1.Calculate the total number of missing values in the dataset 
```{r}
#calculate the total number of missing values 
totMissing <- sum(is.na(data))
totMissing

```
###2.A strategy for filling in all of the missing values in the dataset could use the mean for that 5-min interval. And that's what we do:
```{r}
data$steps_2 <- data$steps   # create a new column of steps derivated from 1st one
nObs <- nrow(data)
iRow <- 1
while(iRow <= nObs) {
  if( is.na(data$steps[iRow]) ) {
    intervalRow <- data$interval[iRow]
    data$steps_2[iRow] = stepsByInterval$mean[which(stepsByInterval$interval==intervalRow)]
  }   
  iRow <- iRow+1
}

```
###3.the new mean and medium of total number of stpes taken per day with missing values
```{r}

by_date2<-group_by(data,date)
stepsByDay2<- summarize(by_date2, total=sum(steps_2))
stepsByDay2
# new mean
meanTotalStepsByDay2<- mean(stepsByDay2$total)
meanTotalStepsByDay2

# new median
medianTotalStepsByDay2<- median(stepsByDay2$total)
medianTotalStepsByDay2
```
###4.Making a histogram of the total number of steps taken each day and calculating and reporting the mean and median total number of steps taken per day.
```{r}
with(stepsByDay2, {    
      barplot(height=stepsByDay2$total, names.arg=date, xlab="date", ylim=c(0,25000),
          ylab="# of steps", main = "Number of steps by day", panel.first=grid())
})
abline(h=meanTotalStepsByDay2, col="blue")
abline(h=medianTotalStepsByDay2, col="red")

```

##Are there difference in activity partterns bewteen weekdays and weekends?
###1.Create a new factor variable in the dataset with two levels ¨C ¡°weekday¡± and ¡°weekend¡± indicating whether a given date is a weekday or weekend day.
```{r}
#distinguish the weekday and weekend using build a week_period function
typeDay<-c("weekend","weekday")
week_period <- function(v_date) {
  i <- as.POSIXlt(as.Date(v_date))$wday
  return( ifelse(i == 0 || i == 6, typeDay[1], typeDay[2]) )
}
# add new variable for data to sign the day type
data$typeDay <- as.factor( sapply(data$date, week_period) )

```
###2.Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
par(mfrow=c(2, 1), mar=c(3, 3, 2, 1), oma=c(1.5, 1, 2, 1))

# make a diagram
for(i in typeDay ){
  stepsByTpDay <- aggregate(steps_2 ~ interval, data = data, subset = (data$typeDay == i), FUN = mean)
  plot(stepsByTpDay, type = "l", main = i, xlab="", ylab="", ylim=c(0,250), xlim=c(0,2400), 
       panel.first=grid(), col.axis='white')
  
  xMarks <- seq(from = 0, to = 2400, by = 100)
  axis(1, at=xMarks, col.axis="blue", las=2, tck=0.02)
  yMarks <- seq(from = 0, to = 250, by = 50)
  axis(2, at=yMarks, col.axis="blue", las=2, tck=0.02)
}

mtext( "interval (5-minute)", side=1, outer=TRUE, col="blue", font=1 )  
mtext( "Mean number of steps", side=2, outer=TRUE, col="blue", font=1, cex=0.9 )  
mtext( "Activity patterns between weekdays and weekends", side=3, outer=TRUE, col="maroon", font=2 )  
box("outer", col="maroon") 

```
